Designing Types for R, Empirically

### 2020-10-14 paper/main.tex
sec 2

The goal of this paper is to gather data that can be used as input to
the process of designing a type system for R.  To focus our work, we
chose to limit the scope of investigation on the design of a language
for function type signatures.  Thus, this paper will use the collected
data to document the signature of user-defined functions.  For this,
we design a simple type language, one that matches the R data types.
We then extract type signature from execution traces of a corpus of
widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We
validate the robustness of the extracted type signature by
implementing a contract checker that weaves the types around their
respective functions, and use a both a large number of clients of the
target packages as well as end-user code for validation.  To sum up,
we make the following contributions:
\begin{itemize}
\item We implemented scalable and robust tooling to automatically
  extract type signatures and instrument R functions with checks
  based on their declared types.
\item We carried out a corpus analysis of \PACKAGES widely used and
  maintained packages to extract function type signatures and
  validated the robustness of the inferred type signatures against
  \NUMFILESRUN programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for R.
\item Our data and code are open source and publicly
  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
  \url{github.com/PRL-PRG/contractr}, {\bf TODO add DATA link}}
\end{itemize}

@@

Dynamic programming languages such as Racket, JavaScript, PHP and Lua
have been extended post factum with static type systems.  In each
case, the type system was carefully engineered to match the salient
characteristics of its host language and to foster a particular
programming style. For example, Typed Racket emphasizes functional
programming and migration from untyped to fully typed
code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
focus on object-oriented features of PHP and JavaScript,
respectively. They allow users to intersperse typed and untyped code
in a fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the
type system tried to account for the myriad ways Lua programmers use
tables.  Other languages adopted a mix of typed and untyped code by
design. In Julia, type annotations are needed for method dispatch and
performance~\cite{oopsla18a}. In Thorn, users could freely move
between typed and untyped code thanks to the addition of like
types~\cite{popl10}. Lastly C\# is an example of a statically typed
language which added a dynamic type~\cite{Bierman10}.

@@

The R Project is a key tool for data analysis.  At the heart of the R
project is a vectorized, dynamic, lazy, functional, object-oriented
language with an unusual combination of features~\cite{ecoop12}.  R
was designed~\citet{R96} as a successor to S~\cite{S88}.

It's basic building blocks are vectors of primitives (primitives in R
are characters (or strings), integers, logicals (or booleans),
doubles, complex numbers, and raw bytes).  To say that R is vectorized
implies that primitive values are {\it always} vectors.  For example,
the literal \code{2.5} is a unit-length vector containing a single
double.  Vectors can be explicitly constructed using the constructor
\code{c(...)}, as in \code{c(1L, 2L, 3L)} which creates a vector of
three integers.

The main object of our work is functions, and R functions have a
number of quirks.  Arguments can be assigned arbitrary expressions as
default values, and functions can support variable
argument lists, and functions can be called by specifying argument
names explicitly in the call.  To illustrate all of this in a single
example, consider:

@@

\noindent
This function has four formal parameters, \k x, \k{...}, \k y and \k
z. Argument \k x can be bound positionally or passed by name. The
vararg argument, \k{...}, is always passed positionally. The remaining
two arguments must be passed by name as they are preceded by the
\k{...} parameter.  Arguments \k y and \k z have default values, in
the case of \k z this is a constant, but \k{y}'s default value is an
expression that depends on the value of \k z (if \k z is not zero, \k
y will default to \k{NULL}). The body of the function will add
parameters \k x and \k y with either the scalar 0 or the result of
concatenating the varargs into a primitive vector. The function
\k{missing} tests if an argument was explicitly passed. The following
are some valid invocations of \k f:

\begin{lstlisting}
 > f(1)    
   [1] 2             # a double vector, y is 0, ... is missing
 > f(1, 2) 
   [1] 4             # a double vector, y is 0, ... is 2
 > f(1, 2, 3)
   [1] 4 5           # a double vector, y is 0, ... is 2, 3
 > f(2, 3, x=1)
   [1] 4 5           # a double vector, y is 0, ... is 2, 3
 > f(x=1, y=1)
   [1] 2             # a double vector, y is 1, ... is missing
 > f(x=1, z=1)
   numeric(0)        # a double vector of length 0, y is NULL
 > f(1L, 2L, y=1L)
   [1] 4             # an integer vector, y is integer 1, ... is integer 2
 > f(1, y=c(1,2))
   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
\end{lstlisting}

\noindent
The above hints at the polymorphism of the language: \k f may return a
vector of integers or of doubles, the length of the vector depends on
the length of the varargs, and of \k x and \k y.  As we mentioned, the
language does not really differentiate between scalar and vectors.
Some more exotic types that can be encountered include vectors of
complex number and list of arbitrary types. Function \k f can be
invoked with those as well.

\begin{lstlisting}
 > c1 <- complex(re=1, im=2)
 > c2 <- complex(re=2, im=1)
 > f(c1)
   [1] 1             # a double vector, x is complex, y is double
 > f(c1, y=c2)
   [1] 2+1i          # a complex vector, x and y are complex
 > l1 <- list(1, 2)
 > l2 <- list(c1, c2)
 > f(l1) 
   [1] 1             # a double vector, x is a list of doubles
 > f(l1, y=l2)
   [[1]]             # a list of complex, x is a list of doubles
   [1] 1+2i          #                      y is a list of complex

   [[2]]
   [1] 2+1i
\end{lstlisting}

R has a builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language; these are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
contexts a \k{logical} will be coerced to \k{integer}, and an \k{integer} will
be coerced to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.

All vectorized data types have a distinguished missing value denoted
by \code{NA} (for ``not available'').  The default type of \code{NA}
is \k{logical}.  We can see that \code{typeof(NA)=="logical"}, but
\k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==}
\k{"double"}.  In addition to \code{NA}, R has a \code{NULL} value,
and to understand why R has both a \code{NULL} value and \code{NA}
values it is useful to remember that R is primarily designed for
statistics and data science.  In experimental data science, it is
useful to have a notion of a ``missing observation'', which is often
NaN (not a number, or similar) in, e.g., MATLAB and Python.  Since
vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it
  each} primitive type.  There single \code{NULL} null-typed value in
the language to encompass all other cases.  In a sense, you can think
of \code{NA} as representing a missing data point, and \code{NULL} as
a missing data {\it set}.

With one exception, all vectorized data types are monomorphic.  The
exception is the \k{list} type which can hold values of any other type
including \k{list}. For all monomorphic data types, attempting to
store a value of a different type will cause a conversion. Either the
value is converted to the type of the vector, or the vector is
converted to the type of the value.

Over the years, programmers have found the need for a richer type structure
and have added {\it attributes}. The best way to think of attributes is as
an optional map from names to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.  The addition of
attributes lets programmers extend the set of types by tagging data with
user-defined attributes. For example, one could define a vector of four
values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
arithmetic functions will treat \k{x} as a 2x2 matrix.

Another attribute that can be set is the \k{class}.  This attribute
can be bound to a list of class names. For instance,
\code{class(x)<-"human"} will set the class of \k{x} to be \k{human}.
The main usage of attributes is for object-oriented programming. There
are three object-orientation frameworks in R: S3, S4, and R5.  The S3
object system support single dispatch on the class of the first
argument of a function, whereas the S4 object system allows multiple
dispatch (on all arguments). R5 allows for users to define objects in
a more imperative style. Some of the most widely used data types
leverage attributes, e.g., data frames and matrices. A data frame, for
instance, is a list of vectors with a class and a column name
attribute, and matrices are vectors with a \k{dims} attribute.

Scalar data types include the distinguished \k{NULL} value, instances
of classes written using the S4 object system, closures, and
environments.  The implementation of R has a number of other types
listed in Figure~\ref{rtypes} for reference.

>>>>

The goal of this paper is to gather data that can be used as input to
the process of designing a type system for R.  To focus our work, we
chose to limit the scope of investigation on the design of a language
for function type signatures.  Thus, this paper will use the collected
data to document the signature of user-defined functions.  For this,
we design a simple type language, one that matches the R data types.
We then extract call and return types from execution traces of a
corpus of widely used libraries, and finally synthesize type
signatures.  This allows us to see how far one can get with a simple
type language and identify limitations of our design. We validate the
robustness of the extracted type signatures by implementing a contract
checker that weaves types around their respective functions, and use
clients of the target packages for validation.  To sum up, we make the
following contributions:
\begin{itemize}
\item We implemented scalable and robust tooling to automatically
  extract type signatures and instrument R functions with checks
  based on their declared types.
\item We carried out a corpus analysis of \PACKAGES widely used and
  maintained packages to synthesize function type signatures and
  validated the robustness of the signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for R.
\item Our data and code are open source and publicly
  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
  \url{github.com/PRL-PRG/contractr}, {\bf TODO add DATA link}}
\end{itemize}

@@

Dynamic programming languages such as Racket, JavaScript, PHP and Lua
have been extended post factum with static type systems.  In each
case, the type system was carefully engineered to match the salient
characteristics of its host language and to foster a particular
programming style. For example, Typed Racket emphasizes functional
programming and migration from untyped to fully typed
code~\cite{tf-popl08}, Hack~\cite{hack13} and TypeScript~\cite{BAT14}
focus on object-oriented features of PHP and JavaScript,
respectively. They allow users to mix typed and untyped code in a
fine-grained manner.  In the case of Lua~\cite{typedlua2014}, the type
system tried to account for the myriad ways Lua programmers use
tables.  Other languages adopted a mix of typed and untyped code by
design. In Julia, type annotations are needed for method dispatch and
performance~\cite{oopsla18a}. In Thorn, users could freely move
between typed and untyped code thanks to the addition of like
types~\cite{popl10}. Lastly C\# is an example of a statically typed
language which added a dynamic type~\cite{Bierman10}.

@@

The R Project is a key tool for data analysis.  At the heart of R is a
vectorized, dynamic, lazy, functional, object-oriented language with
an unusual combination of features~\cite{ecoop12}.  R was designed by
\citet{R96} as a successor to S~\cite{S88}.

R's main data type is the primitive vector.  Vectors care explicitly
constructed by the constructor \code{c(...)}, as in \code{c(1L, 2L,
  3L)} which creates a vector of three integers.  R has a builtin
notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language; these are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in
many contexts a \k{logical} will be coerced to \k{integer}, and an
\k{integer} will be coerced to \k{double}.  Some odd conversions can
occur in corner cases, such as \k{1<"2"} holds and \k{c(1,2)[1.6]}
returns the first element of the vector, as the double is converted to
an integer. R does not distinguish between scalars and vectors (they
are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}.  With the exception of lists,
all vectorized data types are monomorphic.  A \k{list} can hold values
of any other type including \k{list}. For all monomorphic data types,
attempting to store a value of a different type will cause a
conversion. Either the value is converted to the type of the vector,
or the vector is converted to the type of the value.

\noindent
All vectorized data types have a distinguished missing value denoted
by \code{NA} (for ``not available'').  The type of \code{NA} is
\k{logical} (\code{typeof(NA)=="logical"}), but \k{NA} inhabits every
type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.  R also has a
\code{NULL} value.  In data science, it is useful to have a notion of
a ``missing observation'', since vectors are monomorphic there is a
need for a missing value to inhabit each primitive type. In a sense,
\code{NA} represents a missing data point, and \code{NULL} is a
missing data set.

Over time, programmers have found the need for a richer type
structure.  R supports this with {\it attributes}. One may think of
attributes as an optional map from names to values attached to any
built in type.  Attributes are used to encode various type
structures. They are queried with \k{attributes} and \k{class}.  Using
attributes, programmers can extend the set of types by tagging
data. For example, take the vector of four values,
\code{x<-c(1,2,3,4)} and attach attribute \k{dim},
\code{attr(x,"dim")<-c(2,2)}, to treat \k{x} as a 2x2 matrix.

Another attribute that can be set is the \k{class}, this is one way to
use R for object-oriented programming.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"} will set
the class of \k{x} to be \k{human}.  There are three
object-orientation frameworks in R: S3, S4, and R5.  The S3 object
system support single dispatch on the class of the first argument of a
function, whereas the S4 object system allows multiple dispatch. R5
allows for users to define objects in a more imperative style. Some of
the most widely used data types leverage attributes, e.g., data frames
and matrices. A data frame, for instance, is a list of vectors with a
class and a column name attribute, and matrices are vectors with a
 \k{dims} attribute.

R functions have a number of quirks.  Arguments can be assigned
arbitrary expressions as default values; functions take variable
numbers of parameters, and can be called positionally and nominally.
Consider this example:

@@

\noindent
This function has four formal parameters, \k x, dots, \k y and \k
z. Parameter \k x can be bound positionally or passed by name. The
dots, \k{...}, is always positional. The remaining two parameters must
be passed by name as they are preceded by dots; \k y and \k z have
default values, in the case of \k z this is a constant, but \k{y} is
bound to an expression that depends on \k z's value (if \k z is not
zero, \k y defaults to \k{NULL}). The body of the function will add \k
x and \k y tp either 0 or the result of concatenating the dots into a
primitive vector. The function \k{missing} tests if a parameter was
explicitly passed. The following are some valid invocations of \k f:

\begin{lstlisting}
 > f(1)    
   [1] 2             # a double vector, y is 0, ... is missing
 > f(2, 3, x=1)
   [1] 4 5           # a double vector, y is 0, ... is 2, 3
 > f(x=1, y=1)
   [1] 2             # a double vector, y is 1, ... is missing
 > f(x=1, z=1)
   numeric(0)        # a double vector of length 0, y is NULL
 > f(1L, 2L, y=1L)
   [1] 4             # an integer vector, y is integer 1, ... is integer 2
 > f(1, y=c(1,2))
   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
\end{lstlisting}

\noindent
The above hints at polymorphism: \k f may return a vector of integers
or of doubles of length equal to the max length of its arguments.
